Fast inhibitory loop explains functional map development with short-range inhibition.

### 2016-11-25 paper.tex
changes

All previous topologically organized models of cortical competition have either assumed instantaneous synaptic transmission \cite{?,?} or uniform transmission delays across the different intra-cortical connections \cite{?,?}. This is also the case of the analysis performed by Muir et al. \cite{Muir2014} which questions the ability of intra-cortical connections to explain cortical competition. However, a recent study by Ohana et al. \cite{Ohana2012} have revealed a specific pattern of transmission delays between different neural type combinations. Specifically they found that the
excitatory to excitatory connections are slow while the excitatory to inhibitory connections are fast. In this study we explore the 
possibility that this specific transmission delay pattern is the missing link that can explain how short range inhibition can lead 
to effective cortical competition. We employ computational models to show that the fast inhibitory loop allows for the disynptic
inhibition to be effective and consequently provides the effective Mexican-hat like lateral interactions, thus for the first time 
explaining competition in topologically organized cortical networks with no biologically implausible assumptions. To demonstrate the
proposed properties, in this study we focus on models of development of orientation maps, because this is the most well explored
example of competition driven functional cortical organization development, but these results generalize to development of other
cortical topological properties, as well as potentially other competition based computations.

@@

@@

In this article we will proceed through multiple gradually more complex models of orientation development to address several key questions,
eventually demonstrating that the experimentally identified fast inhibitory loop is a satisfactory explanation for how short-range 
inhibition can support development of cortical functional organization. We will use two different computational abstractions 
to explore the questions at hand. In the fist part of the study we will use computational model that is derived from the LISSOM 
family of models \cite{CMVC}. This choice has three advantages. It allows for a very straightforward explanation of why 
fast inhibitory loop enables short-range inhibition to induce competition. It makes our explanation directly comparable to the 
extensive set of published LISSOM family models, and thus demonstrates that the solution proposed here generalizes to range of other 
functional properties. And finally the LISSOM abstraction enables very fast simulations, thus allowing us to perform 
parameter search analysis that would be otherwise computationally prohibitive. However, as we will explain in section \ref{sec:SM4}, some simplifications
made by the LISSOM abstraction, specifically the instantaneous translation of the neuronal input into it's activity, will leave 
certain questions unanswered. These will be addressed in section \ref{sec:SM4} using more detailed rate model framework.

@@

Ohana et al. \cite{Ohana2012} have shown that transmition delays between different types of cortical neurons are not uniform, specifically they found that on average the transmission delays  between excitatory neurons are $\approx$1.4\,ms, from excitatory to inhibitory neurons are $\approx$0.54\,ms, from inhibitory to excitatory neurons are $\approx$0.98\,ms. The sample size for connections
between inhibitory cells in the study was not large enough to be quantitatively reliable. The key observation here is that the 
combined di-synaptic delay from excitatory to inhibitory and inhibitory to excitatory cells is approximately as long as the mono-synaptic delay
of the excitatory to excitatory connections. Drawing from this, the core insight of this study is that under such pattern of delays
the effective inhibitory interactions are well approximated by the convolution of the excitatory and inhibitory connection kernels, thus 
giving the effective inhibition longer range, consequently fulfilling the essential requirement for cortical competition to occur.

For didactic purposes, let us first explore one example set of conditions under which the above statement holds exactly:

@@

\item The sum of the excitatory to inhibitory , and inhibitory to excitatory delays is exactly the same as the excitatory to excitatory delay
\item There is no inhibitory to inhibitory interaction
\item The synaptic inputs into the excitatory and inhibitory neurons are instantaneously translated into their rate response via positive rectified transfer function.
\item The connection kernels of all neurons are Gaussian kernels with spatial constant $\sigma_{e}$ for excitatory neurons and $\sigma_{i}$ for inhibitory neurons.
\item We will disregard long-range (excitatory) connectivity, and assume that both local excitatory and local inhibitory connections have the same extent , thus $\sigma_{e} = \sigma_{i}$ 
\end{enumerate}

@@

This shows that under these specific assumptions the effective inhibitory interactions between excitatory neurons are $\sqrt{2}$ longer than excitatory and thus
follow the mexican-hat like profile required for for lateral cortical competition underlying functional map development. Even though this essential condition of 
cortical competition is fulfilled in this model configuration, the requirement of local inhibition puts important constraints on the extent of the effective lateral
interactions (relative to lateral excitation). It thus remains unclear wether development of high-quality orientation maps as observed experimentally is 
supported under these conditions. Conveniently, the lateral interaction in the LISSOM family of models is governed by the same equations as ? and ?. 
In the following we will use the GCAL model \cite{Stevens2013}, the latest and most robust addition to the LISSOM family of models to demonstrate that the above specific configuration of effective lateral excitation and inhibition permits development of high-quality orientation maps. To facilitate comparison we will use exactly the same model configuration as in Stevens et al. \cite{Stevens2013}, except to three modifications necessitated by the analysis above:

@@

\item The change in point ? will results in change in the balance of overall excitation and inhibition in the model which is critical 
to correction functional development in the model. We will thus modify the strength of the lateral inhibition to compensate for changes in ?.

@@

Furthermore, note that the GCAL model used in figure \ref{fig:figure1} only explicitly models excitatory neurons and assumes both direct excitatory and inhibitory
interactions between them, thus corresponding to equation ?. Even though above we have shown that equation ? is equivalent to equation ? to verify 
correctness of our analysis we have ran a single simulation of the GCAL model corresponding to the parameter combination with the highest map 
quality found in figure \ref{fig:figure1}, but with explicitly simulated inhibitory population. In this model we thus do not model direct inhibitory interactions
between excitatory neurons, but instead add excitatory to inhibitory and inhibitory to excitatory connections of the same extend as those of 
excitatory to excitatory (see the assumption \#5). As we have shown above this model should be mathematically equivalent to the simulations 
figure \ref{fig:figure1}D, which is confirmed in figure \ref{fig:figure1}E that shows that the simulation results in formation of identical orientation map as in the figure \ref{fig:figure1}D.

CORRECT THIS END ?????

@@

In the previous section we have shown that under a set of specific assumptions short-range inhibition can induce 
effective mexican-hat like interaction and thus support development of orientation maps. However not all assumptions we 
made were in line with experimental evidence. In this section we will show that one of these assumptions is not necessary, 
specifically that addition of inhibitory connections does not invalidate results of the previous section.

@@

By explicitly modelling the inhibitory neurons in this model we have replaced a single parameter gowerning the strength of inhibitory
later interactions in the model from previous section with three new parameters that set the strength of excitatory to inhibitory,
inhibitory to excitatory and inhibitory to inhibitory connections (all other parameters remaind the same as the 
in the best parametrization found in previous section). Note that in principle there is redundancy in these parameters as
the overall strength of the projections from inhibitory neurons onto both excitatory and inhibitory populations is scaled by the 
excitatory to inhibitory projections strength. Therefore in figure \ref{fig:figure2} we have systematically varied the strength of the added inhibitory to inhibitory
projection expressed relatively to the strength of the inhibitory to inhibitory connections (which was set to 1), while also varying the 
strength of the excitatory to inhibitory projection. We have  investigated the quality of the orientation 
maps that developed under these different levels of inhibitory to inhibitory interactions. As can be seen, high quality maps 
can develop under the full range of the inhibitory to inhibitory interaction strengths, depending on the overall 
excitatory to inhibitory drive. This shows that inclusion of direct inhibitory to inhibitory interactions does not 
invalidate the results of section \ref{sec:SM3}.

>>>>

All previous topologically organized models of cortical competition have either assumed instantaneous synaptic transmission \cite{Kang2003,Levy2011,VonderMalsburg1973,CMVC} or uniform transmission delays across the different intra-cortical connections \cite{?}. This is also the case of the analysis performed by Muir et al. \cite{Muir2014} which questions the ability of the accepted cortical architecture to explain cortical competition. However, a recent study by Ohana et al. \cite{Ohana2012} have revealed a specific pattern of transmission delays between different neural type combinations. Specifically they found that the excitatory to excitatory connections are slow while the excitatory to inhibitory connections are fast. In this study we explore the 
possibility that this specific transmission delay pattern is the missing link that can explain how short range inhibition can lead 
to effective cortical competition. We employ computational models to show that the fast excitatory-to-inhibitory-to-excitatory pathway allows for the disynptic inhibition to generate the effective Mexican-hat like lateral interactions, thus for the first time 
explaining competition in topologically organized cortical networks with no biologically implausible assumptions. To demonstrate the
proposed properties, in this study we focus on models of development of orientation maps, because this is the most well explored
example of competition driven functional cortical organization development, but these results generalize to development of other
cortical topological properties, as well as potentially other competition based computations.

@@

TODO?????

@@

In this article we will proceed through multiple gradually more complex models of orientation development, addressing several 
of the major issues with modelling of this phenomenam, eventually demonstrating that the experimentally identified fast inhibitory loop is a satisfactory explanation for how short-range inhibition can support development of cortical functional organization. We will use two different computational abstractions to explore the questions at hand. In the fist part of the study we will use computational model that is derived from the LISSOM family of models \cite{CMVC}. This choice has three advantages. It allows for a very straightforward explanation of why 
fast inhibitory loop enables short-range inhibition to induce competition. It makes our explanation directly comparable to the 
extensive set of published LISSOM family models, and thus demonstrates that the solution proposed here generalizes to range of other 
functional properties. And finally the LISSOM abstraction enables very fast simulations, thus allowing us to perform 
parameter search analysis that would be otherwise computationally prohibitive. However, as we will explain further in section \ref{sec:SM4}, some simplifications made by the LISSOM abstraction, specifically the instantaneous translation of the neuronal input into it's activity, will leave 
certain questions unanswered. These will be addressed in section \ref{sec:SM4} using more detailed rate model framework.

@@

Ohana et al. \cite{Ohana2012} have shown that transmition delays between different types of cortical neurons are not uniform, specifically they found that on average the transmission delays  between excitatory neurons are $\approx$1.4\,ms, from excitatory to inhibitory neurons are $\approx$0.5\,ms, and from inhibitory to excitatory neurons are $\approx$0.98\,ms. The sample size for connections
between inhibitory cells in the study was not large enough to be quantitatively reliable. The key observation here is that the 
combined di-synaptic delay from excitatory to inhibitory and inhibitory to excitatory cells is approximately as long as the mono-synaptic delay
of the excitatory to excitatory connections. Drawing from this, the core insight of this study is that under such pattern of delays
the effective inhibitory interactions are well approximated by the convolution of the excitatory and inhibitory connection kernels, as we will
show below. This gives the effective inhibition longer range, consequently fulfilling the essential requirement for cortical competition to occur.
For didactic purposes, let us first explore one specific set of conditions under which the above statement holds exactly, we will explore the 
situation when these conditions are relaxed in subsequent sections:

@@

\item The sum of the excitatory to inhibitory and inhibitory to excitatory delays is exactly the same as the excitatory to excitatory delay
\item There is no inhibitory to inhibitory interaction
\item The synaptic inputs into the excitatory and inhibitory neurons are instantaneously translated into their rate response via positive rectified transfer function.
\item The connection kernels of all neurons are Gaussian kernels with spatial constant $\sigma_{e}$ for excitatory neurons and $\sigma_{i}$ for inhibitory neurons.
\item We will assume only local connectivity (disregarding long-range excitatory connections), and assume that both excitatory and inhibitory have the same extent, thus $\sigma_{e} = \sigma_{i}$ 
\end{enumerate}

@@

This shows that under these specific assumptions the effective inhibitory interactions between excitatory neurons are $\sqrt{2}$ longer than excitatory and thus follow the mexican-hat like profile required for for lateral cortical competition underlying functional map development. Even though this essential condition of cortical competition is fulfilled in this model configuration, important constraints on the extent of the effective lateral
interactions (relative to lateral excitation) remain. It is thus still unclear wether development of high-quality orientation maps as observed experimentally is supported under these conditions. Conveniently, the lateral interaction in the LISSOM family of models is governed by the same equations as ? and ?. In the following we will use the GCAL model \cite{Stevens2013}, the latest and most robust addition to the LISSOM family of models to demonstrate that the above specific configuration of effective lateral excitation and inhibition permits development of high-quality orientation maps. To facilitate comparison we will use exactly the same model configuration as in Stevens et al. \cite{Stevens2013} (see Materials and Methods), except to three modifications necessitated by the analysis above:

@@

\item The change in point ? will results in change in the balance of overall excitation and inhibition in the model which is critical 
for successfull functional development in the model. We will thus modify the strength of the lateral inhibition to compensate for changes in ?.

@@

Furthermore, note that the GCAL model used in figure \ref{fig:figure1} only explicitly models excitatory neurons and assumes both direct excitatory and inhibitory
interactions between them, thus corresponding to equation ?. Even though above we have shown that equation ? is equivalent to equation ? to verify 
correctness of our analysis we have ran a single simulation of the GCAL model corresponding to the parameter combination with the highest map 
quality found in figure \ref{fig:figure1}, but with explicitly simulated inhibitory population (figure \ref{fig:figure1}E). In this model we thus do not model direct inhibitory interactions between excitatory neurons, but instead add excitatory to inhibitory and inhibitory to excitatory connections of the same extend as those of excitatory to excitatory pathway (see the assumption \#5). As we have shown above this model should be mathematically equivalent to the simulations figure \ref{fig:figure1}D. Note, however, that the GCAL simulations represent a discrete approximation in both time and space of equations ? and we thus expect small numerical discrepancies. Indeed, orientation maps shown in figure \ref{fig:figure1}E  are nearly identical to those in the figure \ref{fig:figure1}D, with only minor numerical differences, confirming the validity of our approach.

@@

In the previous section we have shown that under a set of specific assumptions short-range inhibition can induce 
effective mexican-hat like interaction and thus support development of orientation maps. However not all assumptions we 
made were in line with experimental evidence. In this section we will numerically show that one of these assumptions is not necessary, 
specifically that addition of inhibitory connections does not invalidate results of the previous section.

@@

By explicitly modelling the inhibitory neurons in this model we have replaced a single parameter gowerning the strength of inhibitory
later interactions in the model from previous section with three new parameters that set: (1) strength of excitatory to inhibitory,
(2) inhibitory to excitatory and (3) inhibitory to inhibitory connections (all other parameters remaind the same as the 
in the best parametrization found in previous section). Note that in principle there is redundancy in these parameters as
the overall strength of the projections from inhibitory neurons onto both excitatory and inhibitory populations is scaled by the 
excitatory to inhibitory projections strength. Therefore in figure \ref{fig:figure2} we have systematically varied the strength of the added inhibitory to inhibitory
projection expressed relatively to the strength of the inhibitory to inhibitory connections (which was set to 1), while also varying the 
strength of the excitatory to inhibitory projection. We have  investigated the quality of the orientation 
maps that developed under these different levels of inhibitory to inhibitory interactions. As can be seen, high quality maps 
can develop under the full range of the inhibitory to inhibitory interaction strengths, depending on the overall 
excitatory to inhibitory drive. This shows that inclusion of direct inhibitory to inhibitory interactions does not 
invalidate the results of section \ref{sec:SM3}.
